---
title: "Chap4"
output: html_document
---

# Chapter 4. Modeling continuous character evolution on a phylogeny 
---
```{r}
library(ape)
library(ggtree)
library(ggplot2)
library(dplyr)
library(readr)
library(aplot)
library(geiger)

```

### Simulating Brownian motion:

```{r}
# set values for time steps and sigma squared parameter
t<-0:100
sig2<-0.01
## simulate a set of random changes
x<-rnorm(n=length(t)-1,sd=sqrt(sig2))
## compute their cumulative sum
x<-c(0,cumsum(x))
# create a plot with nice margins
par(mar=c(5.1,4.1,2.1,2.1))
plot(t,x,type="l",ylim=c(-2,2),bty="n",
xlab="time",ylab="trait value",las=1,
cex.axis=0.8)
```
You may have noticed that instead of simulating additive Brownian evolution step-by-step, we began by simulating all the steps, then we proceeded to compute the state of the process through time by calculating the cumulative sum from time 0 through t for all possible values of t. This is possible to do because a property of the model is that the distribution of changes under Brownian motion is invariant and does not depend on the state of the chain.


```{r}
# set number of simulations
nsim<-100
# create matrix of random normal deviates
X<-matrix(rnorm(n=nsim*(length(t)-1),sd=sqrt(sig2)),
nsim,length(t)-1)
# calculate the cumulative sum of these deviates
# this is now a simulation of Brownian motion
X<-cbind(rep(0,nsim),t(apply(X,1,cumsum)))
# plot the first one
par(mar=c(5.1,4.1,2.1,2.1))
plot(t,X[1,],ylim=c(-2,2),type="l",bty="n",
xlab="time",ylab="trait value",las=1,
cex.axis=0.8)
# plot the rest
invisible(apply(X[2:nsim,],1,function(x,t) lines(t,x),t=t))
```

In fact, the expected variance under Brownian motion is just σ2 multiplied by t (Felsenstein 1973).
To see this, we can simply take our last set of 100 replicate simulations and, for each value
of t, compute the variance across the entire set of simulations.

```{r}
# calculate variance of columns
v<-apply(X,2,var)
# plot the results
par(mar=c(5.1,4.1,2.1,2.1))
plot(t,v,ylim=c(0,1),type="l",xlab="time",
ylab="variance",bty="n",las=1,
cex.axis=0.8)
lines(t,t*sig2,lwd=3,col=rgb(0,0,0,0.1))
legend("topleft",c("observed variance","expected variance"),
lwd=c(1,3),col=c("black",rgb(0,0,0,0.1)),
bty="n",cex=0.8)

#Likewise, the variance at the end of the simulation should just be σ2 (0.01, in our case)
#multiplied by the total time elapsed (100), or about 1.
# find variance at the end of the simulations
var(X[,length(t)])
```


### Brownianmotion on a phylogeny

```{r}
## load phytools package
library(phytools)
## simulate a tree and Brownian evolution on that
## tree using simBMphylo
object<-simBMphylo(n=6,t=100,sig2=0.01,
fsize=0.8,cex.axis=0.6,cex.lab=0.8,
las=1)
```
since lineages can only begin to differentiate after they’ve diverged, the amount of variation that tends to accrue between species appears to increase as a function of the amount of evolutionary time since they shared common ancestry

properties of Brownian motion:
is a stochastic, directionless process. A single realization of Brownian motion might move up or down (figure 4.1),

In fact, if we conducted many phylogenetic Brownian simulations on a given phylogeny,
instead of just one, we would see that on average, the correlation between related taxa would
be exactly proportional to the fraction of shared history they have in common: in other words,
the time from the root of the tree to their MRCA,5 divided by the total tree length.

A scatter plot matrix in which each x, y coordinate of each point in each panel represents the pair of phenotypic values for a single simulation of the Brownian process in each pair of species corresponding to the panel of our matrix.
If the scatter of points for a hypothetical pair of species i and j is tightly clustered to the 1:1 line, then this indicates that (across many replicates of the evolutionary process) species i and j always tend to evolve similar values for the trait. By contrast, if the scatter of points for species i and j is diffuse and uncorrelated, this indicates that there is no tendency in our simulation for species i and j to evolve similar values of the trait acrossmany instances of Brownian evolution on our tree from figure 4.5.

```{r}
## pull the phylogeny out of the object we simulated
## for figure 4.5 using simBMphylo
tree<-object$tree
## simulate 1000 instance of Brownian evolution on that
## tree
X<-fastBM(tree,nsim=1000)

## set the orientation of the axis labels to be
## horizontal
par(las=1)
## create a scatterplot matrix from our simulated
## data using pairs
pairs(t(X)[,tree$tip.label[6:1]],pch=19,
col=make.transparent("blue",0.05),
cex.axis=0.9)
```
### Fitting a Borwninan model

The Brownian model has two different parameters that we’ll try to estimate.
The first of these is the “instantaneous variance” of the stochastic evolutionary process under
Brownian motion, usually referred to as the evolutionary rate: σ2.

The second of these is the initial state of the process. This corresponds to the ancestral
condition at the root node of our tree, and here we’ll denote this quantity as x0.

To estimate the two parameters of our model from a set of observations for species and our tree, we need a criterion for choosing values of σ2 and x0 that best fit our data.
One strategy is to select the values that maximize the probability of obtaining the data that we’ve observed.
This strategy is called estimation by maximum likelihood and was initially developed by Ronald Fisher near the start of the twentieth century (Fisher 1922; see historical review in Aldrich 1997). 
Maximum likelihood estimators have since been shown to possess many desirable statistical properties, such as consistency, efficiency, and asymptotic unbiasedness (Edwards 1992).7

```{r}
## read bacterial data from file
bacteria.data<-read.csv("data/bac_rates.csv", row.names=1)
head(bacteria.data,3)
```

```{r}
bacteria.tree<-read.tree("data/bac_rates.phy")
print(bacteria.tree,printlen=2)

## graph phylogeny using plotTree
plotTree(bacteria.tree,ftype="i",fsize=0.5,
lwd=1,mar=c(2.1,2.1,0.1,1.1))
## add a horizontal axis to our plot
axis(1,at=seq(0,1,length.out=5),cex.axis=0.8)
```
```{r}
name.check(bacteria.tree,bacteria.data)
genome_size<-bacteria.data[,"Genome_Size_Mb"]
genome_size
names(genome_size)<-rownames(bacteria.data)
head(genome_size)
```

```{r}
## fit Brownian motion model using fitContinuous
fitBM_gs<-fitContinuous(bacteria.tree,genome_size)
fitBM_gs
```

From the summary of our fitted model, we see that the maximum likelihood estimate (MLE) of σ2 is about 25.0 and that the MLE of the root state x0 (here given as z0) is around 1.98

How do we interpret these values? Well, since our genome sizes have been measured in Megabases12 (Mb), a σ2 of 25 indicates that under a Brownian process, we’d expect a large number of independently evolving lineages to accumulate the variance among each other of
25 Mb2 after one unit of time.13

Our MLE of x0 of 1.98 means that the most likely state at the root of the tree, under our
model, is a genome size of 1.98 Mb.

The printout also gives us some information about the model fit to our data: the model likelihood and the Akaike information criterion (AIC and AICc14) values.
These quantities are not particularly interpretable in absolute terms—especially since the branch lengths of our tree are not in any especially meaningful units. They will become much more important when we progress to comparing alternative models of evolution for our data later on in this chapter.

To find the MLEs, fitContinuous uses numerical optimization, and in this case, it tells us how many independent optimization iterations were used (optimization iterations) and how frequently optimization converged on the same, best solution.
Brownian motion is such a simple model that most often this latter quantity will be 1.00, indicating that the optimizer found the same optimal solution in 100 percent of iterations. For more complex models, the fraction will often be lower than 100 percent.15

```{r}
## pull our mutation accumulation rate as a named vector
mutation<-setNames(bacteria.data[,"Accumulation_Rate"],
rownames(bacteria.data))
head(mutation)
```

the very highly skewed distribution that we see for this trait is nonetheless a strong signal that Brownian evolution is very
likely a bad fit to this trait on its original scale. Let’s transform mutation accumulation rate to a log-scale and plot that too.

```{r}
## set up for side-by-side plots
par(mfrow=c(1,2),mar=c(6.1,4.1,2.1,1.1))
## histogram of mutation accumulation rates on original scale
hist(mutation,main="",las=2,xlab="",
cex.axis=0.7,cex.lab=0.9,
breaks=seq(min(mutation),max(mutation),
length.out=12))
mtext("(a)",adj=0,line=1)
mtext("rate",side=1,line=4,cex=0.9)
## histogram of mutation accumulation rates on log scale
ln_mutation<-log(mutation)
hist(ln_mutation,main="",las=2,xlab="",
cex.axis=0.7,cex.lab=0.9,
breaks=seq(min(ln_mutation),max(ln_mutation),
length.out=12))
mtext("(b)",adj=0,line=1)
mtext("ln(rate)",side=1,line=4,cex=0.9)
```

```{r}
## fit Brownian motion model to log(mutation accumulation)
fitBM_ar<-fitContinuous(bacteria.tree,ln_mutation)
fitBM_ar
```


### Phylogenetic signal

phylogenetic signal: which we’d define as the tendency for related species to resemble one another more than expected by chance (Revell et al. 2008).
Two different methods for measuring phylogenetic signal of quantitative characters: Blomberg et al.’s (2003) K and Pagel’s (1999a) ..

#### Blomberg et al.’s (2003) K 
is best summarized as a normalized ratio comparing the variance among clades on the tree to the variance within clades.
If the variance among clades is high (compared to the variance within clades), then phylogenetic signal is said to be high. Conversely, if the variance within clades is high (compared to the variance among clades), then phylogenetic signal will be low.
This ratio is then normalized by dividing it by its expected value under a Brownian evolutionary process.As such, Blomberg et al.’s (2003) K has an expected value of 1.0 under evolution by Brownian motion.

```{r}
phylosig(bacteria.tree, genome_size)
```


```{r}
## test for significant phylogenetic signal using
## Blomberg’s K
K_gs<-phylosig(bacteria.tree,genome_size,
test=TRUE,nsim=10000)
K_gs

## set plot margins and font size
par(cex=0.8,mar=c(5.1,4.1,2.1,2.1))
## plot null-distribution and observed value of K
plot(K_gs,las=1,cex.axis=0.9)
```
This result, and our plot of figure 4.9, tells us that despite being considerably less than 1.0, our observed value of phylogenetic signal, K, is still larger than the value of K we’d expect to find if our data for genome size were random with respect to the phylogeny.
Now let’s repeat exactly the same exercise, but this time using mutation accumulation rate.

```{r}
## test for phylogenetic signal in mutation accumulation
## rate
K_ar<-phylosig(bacteria.tree,ln_mutation,
test=TRUE,nsim=10000)
K_ar

## plot the results
par(cex=0.8,mar=c(5.1,4.1,2.1,2.1))
plot(K_ar,las=1,cex.axis=0.9)
```
Our phylogenetic signal is again very low, and this time we can’t reject the null hypothesis.
This means that our observed value of K for mutation accumulation rate (on a log-scale) is entirely consistent with what we’d expected to obtain by chance if species mutation accumulation rates were randomly arrayed on the phylogeny (figure 4.10).

Another quite reasonable question is whether our measured values of phylogenetic signal— rather than being greater than expected by chance—are nonetheless less than expected under a sensible null model for evolution, such Brownian evolution.
There’s no automated routine in R to run this analysis, so to test this null hypothesis, we’ll have to generate a null distribution of values forK under our null hypothesis: Brownian motion.

```{r}
## simulate 10000 datasets
nullX<-fastBM(bacteria.tree,nsim=10000)
## for each, carry out a test for phylogenetic signal
## and accumulate these into a vector using sapply
nullK<-apply(nullX,2,phylosig,tree=bacteria.tree)
## calculate P-values
Pval_gs<-mean(nullK<=K_gs$K)
Pval_gs

Pval_ar<-mean(nullK<=K_ar$K)
Pval_ar

## set up for side-by-side plots
par(mfrow=c(1,2))
## plot for Genome size
## null distribution
hist(c(nullK,K_gs$K),breaks=30,col="lightgray",
border="lightgray",main="",xlab="K",las=1,
cex.axis=0.7,cex.lab=0.9,ylim=c(0,4000))
## actual value as an arrow
arrows(x0=K_gs$K,y0=par()$usr[4],y1=0,length=0.12,
col=make.transparent("blue",0.5),lwd=2)
text(K_gs$K,0.96*par()$usr[4],
paste("observed value of K (P = ",
round(Pval_gs,4),")",sep=""),
pos=4,cex=0.8)
mtext("(a) - Plot for genome size",line=1,adj=0)
## plot for mutation accumulation rate
## null distribution
hist(c(nullK,K_ar$K),breaks=30,col="lightgray",
border="lightgray",main="",xlab="K",las=1,
cex.axis=0.7,cex.lab=0.9,ylim=c(0,4000))
## actual value as an arrow
arrows(x0=K_ar$K,y0=par()$usr[4],y1=0,length=0.12,
col=make.transparent("blue",0.5),lwd=2)
text(K_ar$K,0.96*par()$usr[4],
paste("observed value of K (P = ",
round(Pval_ar,4),")",sep=""),
pos=4,cex=0.8)
mtext("(b) - Plot for mutation accumulation rate",line=1,adj=0)
```

#### Pagel’s λ

λ is a scaling coefficient for the off-diagonal elements in the expected correlations among species that we learned about in chapter 3 (Pagel 1999a).
Thus, values of λ < 1 correspond to less phylogenetic signal than expected under a Brownian motion model. Unlike K, however, λ is not generally well defined outside of the range of (0,1).
As such, is more appropriate for detecting phylogenetic signal that is lower than expected under Brownian motion than the converse.


```{r}
## compute phylogenetic signal, lambda, for genome size
## and mutation accumulation rate
phylosig(bacteria.tree,genome_size,method="lambda")

phylosig(bacteria.tree,ln_mutation,method="lambda")
```

We see immediately that both traits have lambda estimates that are close to 1; however, λ is
higher for genome size than for mutation accumulation rate.
In some ways, this is similar to our findings for K, in which we showed that genome size had phylogenetic signal similar to the expectation under Brownian motion, while mutation accumulation rate had less.

It’s important to note, though, that we do not expect a one-to-one correspondence between K and λ. To the contrary, we suggest that the two metrics actually measure different aspects of phylogenetic signal and thus we should probably expect them to be different, not the same (also see Boettiger et al. 2012 for additional commentary on λ estimation).
The λ method also allows for a hypothesis test of a null that λ = 0.

```{r}
## test for significant phylogenetic signal, lambda,
## in each of our two traits
lambda_gs<-phylosig(bacteria.tree,genome_size,
method="lambda",test=TRUE)
lambda_gs

lambda_ar<-phylosig(bacteria.tree, ln_mutation,
method="lambda",test=TRUE)
lambda_ar
```

Likelihood surfaces for λ for each trait.

```{r}
## plot the likelihood surfaces
## first set plotting parameters, including subdividing
## our plot area into 1 column and two rows
par(mfrow=c(2,1),mar=c(5.1,4.1,2.1,2.1),
cex=0.8)
## plot the likelihood surfaces of lambda for each of our
## two traits
plot(lambda_gs,las=1,cex.axis=0.9,bty="n",
xlim=c(0,1.1))
mtext("(a) - genome size",line=1,adj=0)
plot(lambda_ar,las=1,cex.axis=0.9,bty="n",
xlim=c(0,1.1))
mtext("(b) - ln mutation rate",line=1,adj=0)
```

This tells us that we can reject a null hypothesis of λ = 0 for both phenotypic traits in our data set.

### “Early burst” (a.k.a. EB) model of Blomberg et al. (2003) 

Under this model, the rate of evolution, σ2, starts with some initial value at the root of the
tree and then declines monotonically through time according to an exponential decay function.

```{r}
## set parameters of the EB process
sig2.0<-1.0
a<-0.04
t<-100
## compute sigmaˆ2 as a function of time under this
## process
sig2.t<-sig2.0*exp(-0.04*0:t)
## graph sigmaˆ2 through time
par(mar=c(5.1,4.1,2.1,2.1))
plot(0:t,sig2.t,type="l",xlab="time",
ylab=expression(sigma^2),bty="l",
las=1,cex.axis=0.9)
```
In particular, a declining rate of evolution through timewill tend to result in large differences
between clades and relatively small differences within them.
Just as we did for Brownian evolution earlier in the chapter, we can also graph evolutionary
change through time on a phylogeny under EB using simBMphylo as follows



```{r}
## visualize early-burst evolution using simBMphylo
object<-simBMphylo(6,100,sig2=sig2.t[1:100],
fsize=0.8,cex.axis=0.6,cex.lab=0.8,
las=1)
```

The EBmodel is one inwhich differences between lineages tend to accumulate rapidly near the beginning of a diversification. As such, it has often been linked to the concept of adaptive radiation.

### The Ornstein–Uhlenbeck (OU) model

The final model we’ll consider for this chapter is the Ornstein–Uhlenbeck model of Hansen (1997).
This model is a relatively simple extension of Brownian motion, with an additional parameter (α) that describes the tendency to return toward a particular central value (θ).

Because this model involves evolutionary change toward a particular value, the Ornstein–Uhlenbeck model is most often interpreted as a model for adaptive evolution in which α corresponds to the strength of natural selection and θ to the position of the optimum.


## Fitting and comparative alternative continuous character models

```{r}
## fit EB model to genome size
fitEB_gs<-fitContinuous(bacteria.tree,genome_size,model="EB")
fitEB_gs
```

Our printed model summary tells us that our ML estimated value of the decay parameter, a, is almost exactly zero.
When a = 0.0, the EB model reduces to a constant rate Brownian motion. This is not an error! It merely indicates that when we impose an EB model on our data, the best-fitting EB model is Brownian evolution!

```{r}
## fit OU model to genome size
fitOU_gs<-fitContinuous(bacteria.tree,genome_size,
model="OU")
fitOU_gs
```

In this case, our best-fitting OU model is not a Brownian model (which would correspond
to α = 0.0); α is quite different from zero, in fact.
In this case, the warningmessage we received is very important because it indicates not that
the MLE of α is near 0 (which would be equivalent to a Brownian motion model) but that it
is at the default upper bound for optimization. Let’s change these default bounds and see what
result we obtain.

```{r}
## fit OU model to genome size
fitOU_gs<-fitContinuous(bacteria.tree,genome_size,
model="OU",bounds=list(alpha=c(0,10)))
fitOU_gs
```
This time, there’s no warning message—and both our optimized value of α and our loglikelihood
are higher, which makes sense.
To compare either EB and Brownian or OU and Brownian, we could use a likelihood ratio
test.26 However, to compare all threemodels simultaneously, our best option is probably to use
an information criterion such as the AIC.27
Let’s do just that.

```{r}
## accumulate AIC scores from our three models into
## a vector
aic_gs<-setNames(c(AIC(fitBM_gs),
AIC(fitEB_gs),AIC(fitOU_gs)),
c("BM","EB","OU"))
aic_gs
```

Since our preferred model under this criterion should be the one with the lowest AIC, this
result tells us that the best-supported model (among those tested) for genome size is Ornstein–
Uhlenbeck.
We can also compute Akaike weights28 using the phytools function aic.w.

```{r}
aic.w(aic_gs)
```

Here you can see that the overwhelming majority of weight falls on the OU model—but evidence in support of the different models is somewhat split, and there is also quite a bit of weight on Brownian motion.

Let’s now repeat all of this analysis, but formutation accumulation rate. In this case, we tried
it ahead of time and know the MLE of α for our OU model is going to fall well outside of our
default bounds, so we set bounds=list(alpha=c(0,100)) to see if that helps.

```{r}
## fit EB model
fitEB_ar<-fitContinuous(bacteria.tree,ln_mutation, model="EB")
## fit OU model
fitOU_ar<-fitContinuous(bacteria.tree,ln_mutation,
model="OU",bounds=list(alpha=c(0,100)))
## accumulate AIC scores in a vector
aic_ar<-setNames(c(AIC(fitBM_ar),
AIC(fitEB_ar),AIC(fitOU_ar)),
c("BM","EB","OU"))
## compute and print Akaike weights
aic.w(aic_ar)
fitOU_ar
```

This tells us that a stochastic process with a tendency to revert toward a central value is
better at explaining our data for mutation accumulation rates than is a model of constant-rate
Brownian motion, or a model of early burst evolution, in which large differences between clades
arise early in diversification and then the rate of evolution slows through time. This is actually
quite a common pattern across comparative data sets (Harmon et al. 2010).




