---
title: "Chap1"
output: html_document
---

```{r, warning=FALSE}
library(ape)
library(phytools)
```

Reading a simple Newick text string:
"is a simple way to encode a phylogenetic tree using a series of nested parentheses.More closely nested species are more closely related. For instance, the simple Newick tree ((chimp,human),gorilla); tells us that the operational taxa chimp
and human aremore closely related to each other than either is to gorilla. There are other ways that phylogenetic
trees can be represented in machine-readable text, but the Newick string is by far the most common."

```{r}
text.string<-
"(((((Robin,Iguana),((((Cow,Whale),Pig),Bat),
(Lemur,Human))),Coelacanth),Goldfish),Shark);"
vert.tree<-read.tree(text=text.string)
```

```{r}
#ploting the tree

plot(vert.tree,no.margin=TRUE)

#if you wanna see the arguments
#args(plot.phylo)
```
```{r}
#Plotting the tree with different styles

par(mfrow=c(2,2),mar=c(1.1,1.1,3.1,1.1)) #divide our plotting device into four subplots for a 2 × 2 grid—
#The order of this vector is bottom, left, top, and right—

#plot in (a) using a retangular cladrogram style
plot(vert.tree)
mtext("(a)",line=1,adj=0)

#plot in (b) with a slanted cladogram style
plot(vert.tree,type="cladogram")
mtext("(b)",line=1,adj=0)

#plot in (c) with a unrooted tree syle
plot(unroot(vert.tree),type="unrooted",
lab4ut="axial",x.lim=c(-2,6.5),
y.lim=c(-3,7.5))
mtext("(c)",line=1,adj=0)
```

The "vert.tree" object is a list containing (at least) the following elements:
1. edge: a 20 × 2 (in this case) matrix containing starting and ending indices for the nodes
subtending each branch of the phylogeny. By convention, tip nodes (that is, those
corresponding to species or operational taxa) are numbered 1 through N for N species,
while internal nodes are numbered N +1 (at the root) through N+ the number of internal
nodes
2. Nnode: an integer value giving the total number of internal nodes in the tree.
3. tip.label: a character vector of length N containing the labels for all the tips or
terminal taxa in the phylogeny.

We can visualize this structure like this:

```{r}
plotTree(vert.tree,offset=1,type="cladogram")
labelnodes(1:(Ntip(vert.tree)+vert.tree$Nnode), 1:(Ntip(vert.tree)+vert.tree$Nnode),interactive=FALSE,cex=0.8)
```

