---
title: "Chap2"
output: html_document
---

## Chapter 2 - Phylogenetically independent contrasts

```{r, warning=FALSE}
library(ape)
library(phytools)
library(readr)
library(geiger)
```

Hypotheses about evolutionary correlations can be tested using a method known as phylogenetic independent contrasts (Felsenstein 1985).

In this American Naturalist article, entitled “Phylogenies and the Comparative Method,” Felsenstein (1985) fundamentally and irreversibly changed the way in which the field of comparative biology collectively viewed the analysis of species data. He did this in large part by illustratingwhy data points obtained from species related to one another by a phylogeny could not and should not be treated as independent from the point of view of traditional statistical analysis

### 2.2.1 Felsenstein’s worst-case scenario

In particular, Felsenstein (1985) used a kind of worst case of sorts3 to cleverly illustrate how phylogenetic nonindependence of species could lead to vast overconfidence in the inference of an evolutionary correlation between traits—if this correlation were to be estimated without taking the phylogeny properly into consideration.
The following code (and Figure 2.1) illustrates Felsenstein’s worst-case scenario using simulated data.

3“A worst case of sorts for the naive analysis is shown . . . where the phylogeny shows that a large number of species actually consist of two groups of moderately close relatives. . . . There appears to be a significant regression of Y on X. If the points are distinguished according to which monophyletic group they came from. . . we can see that there are two clusters. Within each of these groups there is no significant regression of one character on the other. The means of the two groups differ, but since there are only two group means they must perforce lie on a straight line, so that the between-group regression has no degrees of freedom and cannot be significant. Yet a regression assuming independence of the species finds a significant slope (P<.05). It can be shown that there are more nearly 3 than 40 independent points in the diagram” (Felsenstein 1985, p. 4).


```{r}
set.seed(1)
## read in tree
tree<-read.tree(
text="((A,B,C,D,E,F,G,H,I,J,K,L,M),
(N,O,P,Q,R,S,T,U,V,W,X,Y,Z));")
## set branch lengths on the tree
tree<-compute.brlen(tree,power=1.8)
## simulate data, independently for x & y
x<-fastBM(tree)
y<-fastBM(tree)

## Calculate correlation and p-value for the entire tree (both clades together)
cor_full_test <- cor.test(x, y)

## Calculate correlation and p-value for clade A (1 to 13)
cor_clade_A_test <- cor.test(x[1:13], y[1:13])

## Calculate correlation and p-value for clade B (14 to 26)
cor_clade_B_test <- cor.test(x[14:26], y[14:26])

## Print correlations and p-values
cat("Correlation for full tree: ", cor_full_test$estimate, "p-value:",cor_full_test$p.value, "\n")
cat("Correlation for clade A: ", cor_clade_A_test$estimate, "p-value:",cor_clade_A_test$p.value, "\n")
cat("Correlation for clade B: ", cor_clade_B_test$estimate, "p-value:",cor_clade_B_test$p.value, "\n")
```

```{r}
## plot the results with clades A & B labeled
## split plotting area
par(mfrow=c(1,3))
plot(x,y,bty="n",las=1)
points(x[1:26],y[1:26],pch=21,bg="grey",
cex=1)
## Fit a linear model for the regression line for the full data
lm_full <- lm(y ~ x)
abline(lm_full, col = "black", lwd = 2)  # Add regression line

## Add regression equation and R-squared value to the plot
text(x = min(x) + 1, y = max(y) - 0.5, 
     labels = paste("R² = ", round(summary(lm_full)$r.squared, 2), 
                    "\np = ", round(coef(summary(lm_full))[2,4], 5),
                    "\n corr: ", round(cor_full_test$estimate,2), ", p =",round(cor_full_test$p.value,3)),
     cex = 1, col = "black")

mtext("(a)",line=1,adj=0,cex=1.5)
## graph tree
plotTree(tree,type="cladogram",ftype="off",
mar=c(5.1,4.1,3.1,2.1),color="darkgray",
xlim=c(0,1.3),ylim=c(1,Ntip(tree)))
## add points at the tips of the tree to match those
## on our scatterplot
points(rep(1,13),1:13,pch=21,bg="red",
cex=1.2)
points(rep(1,13),14:26,pch=22,bg="blue",cex=1.2)
## add clade labels to the tree
cladelabels(tree,"A",node=28,offset=2)
cladelabels(tree,"B",node=29,offset=2)
mtext("(b)",line=1,adj=0,cex=1)
## create scatterplot of x & y
par(mar=c(5.1,4.1,3.1,2.1))
plot(x,y,bty="n",las=1)
points(x[1:13],y[1:13],pch=21,bg="red",
cex=1.2)
points(x[14:26],y[14:26],pch=22,bg="blue",cex=1.2)


lm_red <- lm(y[1:13] ~ x[1:13])
abline(lm_red, col = "red", lwd = 2)  # Add regression line

## Add regression equation and R-squared value to the plot
text(x = min(x) + 2, y = max(y) - 2.5, 
     labels = paste("R² = ", round(summary(lm_red)$r.squared, 2), 
                    "\np = ", round(coef(summary(lm_red))[2,4], 5),
                    "\n corr: ", round(cor_clade_A_test$estimate,2), ", p =",round(cor_clade_A_test$p.value,3)),
     cex = 1, col = "red")


lm_blue <- lm(y[14:26] ~ x[14:26])
abline(lm_blue, col = "blue", lwd = 2)  # Add regression line

## Add regression equation and R-squared value to the plot
text(x = min(x) + 1.3, y = max(y) - 0.3, 
     labels = paste("R² = ", round(summary(lm_blue)$r.squared, 2), 
                    "\np = ", round(coef(summary(lm_blue))[2,4], 5),
                    "\n corr: ", round(cor_clade_B_test$estimate,2), ", p =",round(cor_clade_B_test$p.value,3)),
     cex = 1, col = "blue")

mtext("(c)",line=1,adj=0,cex=1)
```

Prior to the foundational contribution of Felsenstein (1985), it would not have been uncommon to interpret a simple correlation between two variables as evidence that they had coevolved. What Felsenstein’s paper served to illustrate was that just such a pattern can easily arise in phylogenetic data even absent a genuine evolutionary relationship between traits. Felsenstein’s (1985) article, however, was much more than a critique of the existing paradigm. It also presented a solution for correlation or regression analysis of species data. This was a specific type of data transformation that has come to be known as phylogenetically independent contrasts or PICs. Felsenstein’s PIC method is the subject of the present chapter. This method provides a way to test evolutionary correlations among characters, as we’ll illustrate by example, below.

The principle behind Felsenstein’s method is both remarkably elegant and incredibly simple. Felsenstein (1985) merely pointed out that while species data per se are not independent (owing to shared history due to the phylogeny), the differences (or contrasts) between species are. Likewise, the differences (if properly corrected) between trait values that have been interpolated for internal nodes are also independent. Thus, for a bifurcating N-taxon species tree consisting of observations for two or more continuous traits, one can compute a total of N −1 phylogenetically independent contrasts for each trait. These contrasts, once normalized, can then be used in standard regression or correlation analysis.

Let's use our primate dataset again

## Primate dataset

```{r}
primate_tree <- read.nexus("primates_data/consensusTree_10kTrees_Primates_Version3.nex")

primate_data <- read_csv("primates_data/primates_data_isler_vanschaik.csv")
primate_data <- as.data.frame(primate_data)
colnames(primate_data)[colnames(primate_data) == "Maximum lifespan [y]"] <- "maxlongevity_y"
colnames(primate_data)[colnames(primate_data) == "Gregariousness"] <- "gregariousness"
colnames(primate_data)[colnames(primate_data) == "Allonursing"] <- "allonursing"
colnames(primate_data)[colnames(primate_data) == "Communal nesting, babysitting, retrieval"] <- "communal_nesting_babysitting"
colnames(primate_data)[colnames(primate_data) == "Protection"] <- "protection"
colnames(primate_data)[colnames(primate_data) == "Body mass [g]"] <- "body_mass"
colnames(primate_data)[colnames(primate_data) == "Age at first reproduction [y]"] <- "reproduction_age_y"
primate_data$species <- gsub("_", "_", primate_data$Genus_species)
primate_df <- subset(primate_data, select = c(species,body_mass,maxlongevity_y,reproduction_age_y,gregariousness,allonursing,communal_nesting_babysitting,protection,provisioning_Male,provisioning_others,carryng_Male,carrying_others))
primate_df$species <- tolower(primate_df$species)
primate_tree$tip.label <- tolower(primate_tree$tip.label)
rownames(primate_df) <- primate_df$species
primate_df <- primate_df %>% filter(!is.na(maxlongevity_y))

chk <- name.check(phy = primate_tree, data = primate_df)

tree<-drop.tip(primate_tree,chk$tree_not_data)

# Prune tree to retain only matching tips
pruned_data<-primate_df[pruned_tree$tip.label,]
pruned_data$body_mass <- as.numeric(pruned_data$body_mass)
pruned_data$sociality_total <- pruned_data$gregariousness + pruned_data$allonursing + pruned_data$protection + pruned_data$communal_nesting_babysitting+pruned_data$carrying_others +pruned_data$provisioning_others
pruned_data$sociality_interaction_total <- 1+pruned_data$gregariousness*pruned_data$communal_nesting_babysitting*(pruned_data$allonursing + pruned_data$protection + pruned_data$provisioning_others + pruned_data$carrying_others +pruned_data$carryng_Male +pruned_data$provisioning_Male)
#,"gregariousness","allonursing","communal_nesting_babysitting","protection","provisioning_Male","provisioning_others","carryng_Male","carrying_others"
df <- pruned_data[, c("body_mass","maxlongevity_y","sociality_total","sociality_interaction_total")]
rownames(df) <- pruned_data$species
head(df)

```
```{r}
## set margins of the plot
par(mar=c(5.1,5.1,1.1,1.1))
## create scatterplot
plot(sociality_interaction_total~body_mass,data=df,
xlab="body mass (g)",
ylab=expression(paste("sociality (gregariouness and alloparenting")),
pch=21,bg="gray",cex=1.2,log="xy",las=1,cex.axis=0.7,
cex.lab=0.9,bty="n")
```

### Modeling

```{r}
fit.ols<-lm(log(sociality_interaction_total)~log(body_mass),data=df)
fit.ols
```

```{r}
summary(fit.ols)
```

```{r}
## set margins and graph scatterplot
par(mar=c(5.1,5.1,1.1,1.1))
plot(sociality_interaction_total~body_mass,data=df,
xlab="body mass (kg)",
ylab=expression(paste("sociality (gregariouness and alloparenting")),
pch=21,bg="gray",cex=1.2,log="xy",las=1,
cex.axis=0.7,cex.lab=0.9,bty="n")
## add the line of best fit from lm
lines(df$body_mass,exp(predict(fit.ols)),lwd=2,col="darkgray")
```

But wait! What about the phylogeny?
A key assumption of linear regression analysis is that the residual error in the model is
independently and identically distributed.17

```{r}
## plot phylogeny of mammals
plotTree(tree,ftype="i",fsize=0.7,lwd=1)
## add node labels to the plotted tree
nodelabels(bg="white",cex=0.5,frame="circle")
```

The function to compute contrasts is fromthe ape package and is called pic. Let’s go ahead and calculate contrasts for home range and for body size.
We’re going to do this in two steps. First, we’ll pull out separate vectors for each character and assign the vector names that correspond to the row names of the data frame mammalHR using the handy function setNames. Next, we’ll proceed to compute contrasts for each of these new, named vectors using the ape package PIC function, called pic.



```{r}
## pull our home range and body mass as
## numeric vectors
sociality<-setNames(df[,"sociality_interaction_total"],
rownames(df))
bodyMass<-setNames(df[,"body_mass"],
rownames(df))
longevity<-setNames(df[,"maxlongevity_y"],
rownames(df))
## compute PICs for home range and body size

#tree <- root(tree, resolve.root = TRUE)
tree <- multi2di(tree)

pic.sociality<-pic(log(sociality),tree)
pic.bodymass<-pic(log(bodyMass),tree)
pic.longevity<-pic(log(longevity),tree)

head(pic.sociality,n=20)
```

We can immediately see that the contrasts take bothpositive andnegative values. This makes sense, because contrasts are (standardized21) differences in trait values between sister species or nodes. These differences can obviously be positive if, say, the right daughter node has a higher value for the trait than the left daughter node or negative if the opposite is true. In addition, if we compare either of our PIC vectors to the tree we plotted in figure 2.4,
we should also see that our vector has names that correspond to the node indices of the tree. This toomakes sense, because in a bifurcating22 tree, each contrast subtends one and only one node.

Now we’re ready to fit our linear model.We need keep in mind, however, that we’ll need to fit
this regression model through zero—that is, without an intercept term. This is because for any node in the phylogeny, the rotation of the right and left daughters of that node is arbitrary—and, as such, so is the direction of subtraction of the contrasts.23 Our linear model for those contrasts should thus go through the point (0,0) on our plot (Felsenstein 1985).
To do that in R, we can either append +0 or -1 to our formula in lm.24 Let’s use +0.

```{r}
## fit linear model to PICs without intercept
fit.pic<-lm(pic.sociality~pic.bodymass+0)
fit.pic
```

```{r}
summary(fit.pic)
```

```{r}
x = log(df$body_mass)
y = log(1 + df$sociality_total)
## graph phylomorphospace projection
phylomorphospace(tree,cbind(x,y),label="off",
node.size=c(0,0),bty="n",las=1)
## overlay points onto the phylomorphospace plot
points(x,y,pch=21,bg="gray",cex=1.2)
## add gridlines
grid()
## clip plot
clip(min(x),max(x),min(y),max(y))
## add fitted regression line
abline(fit.ols,lwd=2)
```



```{r}
## set margins
par(mar=c(5.1,5.1,1.1,1.1))
## graph scatterplot of contrasts
plot(pic.sociality~pic.bodymass,
xlab="PICs for log(body mass)",
ylab="PICs for log(sociality)",
pch=21,bg="gray",cex=1.2,las=1,
cex.axis=0.7,cex.lab=0.9,bty="n")
## add gridlines to the plot
abline(h=0,lty="dotted")
abline(v=0,lty="dotted")
## reset graphing limits of the plot to the
## x/y range of our PICs
clip(min(pic.bodymass),max(pic.bodymass),
min(pic.sociality),max(pic.sociality))
## graph our fitted line
abline(fit.pic,lwd=2,col="darkgray")
```

The highly significant correlation that we measure in the OLS regression analysis was a spurious relationship driven by the phylogeny.
A useful way to think about what we found is the x and y seem to be correlated but not evolutionarily correlated. That is to say, the OLS result tells us that we can use x to predict y— but only because (by knowing its value for x) we know something about where that species occurs in the tree, and so we can predict something about y. On the other hand, the nonsignificant linear regression of the contrasts tells us that40 we
find no evidence of an evolutionary tendency for x and y to coevolve.


Now, let's check longevity:

```{r}
fit_longevity.pic<-lm(pic.sociality~pic.longevity+0)
fit_longevity.pic
summary(fit_longevity.pic)

## set margins
par(mar=c(5.1,5.1,1.1,1.1))
## graph scatterplot of contrasts
plot(pic.sociality~pic.longevity,
xlab="PICs for log(longevity)",
ylab="PICs for log(sociality)",
pch=21,bg="gray",cex=1.2,las=1,
cex.axis=0.7,cex.lab=0.9,bty="n")
## add gridlines to the plot
abline(h=0,lty="dotted")
abline(v=0,lty="dotted")
## reset graphing limits of the plot to the
## x/y range of our PICs
clip(min(pic.longevity),max(pic.longevity),
min(pic.sociality),max(pic.sociality))
## graph our fitted line
abline(fit.pic,lwd=2,col="darkgray")
```

